A Triple Origin for Twin Blue Stragglers in Close Binaries

### 2019-01-14 BSS7782.tex
updated some figures

\title{A Triple Origin for Twin Blue Stragglers in Compact Binaries}

@@

We propose a formation mechanism for twin blue stragglers in compact
binaries that involves mass transfer from an evolved outer tertiary
companion on to the inner binary via a circumbinary disk.  We apply
our scenario to the observed double blue straggler system Binary 7782
in the old open cluster NGC 188, and show that its observed properties
are naturally reproduced within the context our proposed model.
Within the context of the hypothetical formation mechanism proposed
here, the presented work predicts the following properties for the
post-mass transfer double BS tertiary: (1) For the outer tertiary
orbit, the initial orbital period should lie between 220 days
$\lesssim$ P$_{\rm out}$ $\lesssim$ 1100 days \simon{Do you think that
  the initial binary could be as large as 1400 days?}, assuming
initial masses for the inner binary components of $m_{\rm 1} = 1.1$
M$_{\odot}$ and $m_{\rm 2} =$ 0.9 M$_{\odot}$ and an outer tertiary
mass of $m_{\rm 3} = 1.4$ M$_{\odot}$.  After Roche-lobe overflow, the
outer star turns into a white dwarf (WD). There is a correlation
bewteen the mass of this WD and the outer orbital period in the sense
that more massive white dwarfs will be on wider orbits.  (3) The
rotational axes of the BSs in the inner binary should be aligned with
each other and the orbital plane of the outer tertiary WD. (4) The BSs
in the inner binary should have roughly equal masses, independent of
their initial masses.  This predicts that the initially lower mass MS
star should accrete the most, and should hence be enriched more
effectively by the accreted material.  This could be observable in the
surface layers of a radiative star (i.e., He, C and O enrichment if
the donor is an RGB star, and/or s-process elements if the donor is an
AGB star). (5) Relative to old dense clusters with high velocity dispersions, 
twin BSs in compact binaries formed from the proposed 
mechanism should be more frequent in the Galactic field and younger 
open clusters with ages $\lesssim$ 4-6 Gyr, since then the donor will
have a radiative envelope. (6) the orbit of the binary BS will be
compact (typically $\aplt 0.3$\,au) and close to circular ($e \aplt
0.2$).

@@

Blue straggler (BS) stars are brighter and bluer than the
main-sequence turn-off (MSTO) in a cluster colour-magnitude diagram
\citep[e.g.][]{1953AJ.....58...61S,2014ApJ...782...49S}.  Two
primary channels for BS formation have been proposed: mass transfer
from an evolved donor on to a main-sequence star in a binary star
system
\citep[e.g.][]{1964MNRAS.128..147M,1997A&A...328..143P,2009Natur.457..288K,2011MNRAS.410.2370L,2011Natur.478..356G},
and direct stellar collisions involving main-sequence stars possibly
mediated via binaries
\citep[e.g.][]{hills75,1997A&A...328..130P,2007ApJ...661..210L,2013MNRAS.428..897L,2013MNRAS.429.1221H,2018arXiv181100058P}.
Other possible, albeit related, formation mechanism includes mergers
of close main-sequence binaries (MS-MS) \cite{2018arXiv181100058P},
and mergers of the inner binaries of hierarchical triple star systems
induced by Lidov-Kozai oscillations coupled with tidal damping
\citep[e.g.][]{2009ApJ...697.1048P}.

In spite of these specific predictions for the expected properties of
BSs formed from each of the above production mechanisms, there exist
many BSs with observed properties that defy these simple scenarios.
For example, in the old open cluster M67, there lurks a candidate
triple system that is posited to host two BSs
\citep{2001A&A...375..375V,2003AJ....125..810S}.  The observations suggest that the
outer tertiary is itself a BS, with a mass $\sim$ 1.7 M$_{\odot}$ and
orbiting the inner binary with a period of 1188.5 days
\citep{{2003AJ....125..810S}.  The inner binary has a period of only 1.068 days
\citep{2001A&A...375..375V}, and hosts a BS of mass 2.52 M$_{\odot}$.  In
order to form this system \del{after convolving its observed
  properties with its host cluster age,} we require at least five
stars in order to conserve the total system mass \citep{2011MNRAS.410.2370L}.  As
explained in \citet{2011MNRAS.410.2370L}, this is strongly indicative of a
dynamical origin for the system, and a single direct interaction
involving a binary and a triple that resulted in two separate
collision events is the most probable explanation for its origin
(i.e., a single interaction involving two multiples with two or more
stars is a more likely scenario to produce this system than two
back-to-back direct binary-binary interactions)
\citep{2004MNRAS.350..615G}.  \simon{Maybe add something about the
  curious binary AE Aurigae? - NL2: For sure.  I still don't know anything about it... But if I get time over the next few days I will do my research and write something.}

@@

\begin{figure}[ht!]
  \includegraphics[width=\columnwidth]{fig_mass_vs_semimajor_axis.pdf}
  \caption{Evolution of the total mass of the inner binary as a
    function of the orbital separation for four calculations with
    somewhat different initial conditions (see legend).  The initial
    binary for the orange curve is presented in
    fig.\,\ref{fig:topview_at_t0}, and in
    fig.\,\ref{fig:topview_at_t1000day} we present the final
    conditions.
\label{fig:mass_vs_semimajor_axis}}
\end{figure}

>>>>

\title{A Triple Origin for Twin Blue Stragglers in Close Binaries}

@@

We propose a formation mechanism for twin blue stragglers in compact
binaries that involves mass transfer from an evolved outer tertiary
companion on to the inner binary via a circumbinary disk.  We apply
this scenario to the observed double blue straggler system Binary 7782
in the old open cluster NGC 188, and show that its observed properties
are naturally reproduced within the context of the proposed model.
Based on this model, we predict the following properties for twin blue
stragglers: (1) For the outer tertiary orbit, the initial orbital
period should lie between 220 days $\lesssim$ P$_{\rm out}$ $\lesssim$
1100 days, assuming initial masses for the inner binary components of
$m_{\rm 1} = 1.1$ M$_{\odot}$ and $m_{\rm 2} =$ 0.9 M$_{\odot}$ and an
outer tertiary mass of $m_{\rm 3} = 1.4$ M$_{\odot}$.  After
Roche-lobe overflow, the outer star turns into a white dwarf (WD) of
0.43 to 0.54\,\MSun. There is a correlation between the mass of this
WD and the outer orbital period in the sense that more massive white
dwarfs will be on wider orbits.  (3) The rotational axes of both the
BSs in the inner binary will be aligned with each other and the
orbital plane of the outer tertiary WD. (4) The BSs in the inner
binary will have roughly the same masses, independent of their initial
masses.  This follows from the more copious accretion of mass onto the
initially lower mass star.  This star should therefore be enriched
more effectively by the accreted material.  As a result, one of the
blue stragglers will appear to be more enriched by He, C and O
if the donor started to overflow its Roche lobe at the red
giant branch, or with s-process elements if the donor was on the
asymptotic giant branch.  (5) Relative to old dense clusters with high
velocity dispersions, twin BSs in close binaries formed from the
proposed mechanism should be more frequent in the Galactic field and
younger open clusters with ages $\lesssim$ 4-6 Gyr, since then the
donor will have a radiative envelope. (6) the orbit of the binary BS
will be close (typically $\aplt 0.3$\,au) and close to circular ($e
\aplt 0.2$).

@@

Blue straggler (BS) stars are brighter and bluer than the
main-sequence turn-off in a cluster colour-magnitude diagram
\citep[e.g.][]{1953AJ.....58...61S,2014ApJ...782...49S}.  Two
primary channels for BS formation have been proposed: mass transfer
from an evolved donor on to a main-sequence star in a binary star
system
\citep[e.g.][]{1964MNRAS.128..147M,1997A&A...328..143P,2009Natur.457..288K,2011MNRAS.410.2370L,2011Natur.478..356G},
and direct stellar collisions involving main-sequence stars possibly
mediated via binaries
\citep[e.g.][]{hills75,1997A&A...328..130P,2007ApJ...661..210L,2013MNRAS.428..897L,2013MNRAS.429.1221H,2018arXiv181100058P}.
Other possible, albeit related, formation mechanism includes mergers
of close main-sequence binaries (MS-MS) \cite{2018arXiv181100058P},
and mergers of the inner binaries of hierarchical triple star systems
induced by Lidov-Kozai oscillations coupled with tidal damping
\citep[e.g.][]{2009ApJ...697.1048P}.

In spite of these specific predictions for the expected properties of
BSs formed from each of the above production mechanisms, many BSs
exist with observed properties that defy these simple scenarios.  For
example, in the old open cluster M67, there lurks a candidate triple
system that is posited to host two BSs
\citep{2001A&A...375..375V,2003AJ....125..810S}.  The observations
suggest that the outer tertiary is itself a BS, with a mass $\sim$ 1.7
M$_{\odot}$ and orbiting the inner binary with a period of $\sim
1188.5$ days \citep{2003AJ....125..810S}.  The inner binary has a
period of only $\sim 1.068$ days \citep{2001A&A...375..375V}, and
hosts a BS of mass $\sim 2.52$ M$_{\odot}$.  In order to form this
system we require at least five stars in order to conserve the total
mass in the system \citep{2011MNRAS.410.2370L}.  As explained in
\citet{2011MNRAS.410.2370L}, this is strongly indicative of a
dynamical origin for the system, and a single direct interaction
involving a binary and a triple that resulted in two separate
collision events is the most probable explanation for its origin
(i.e., a single interaction involving two multiples with two or more
stars is a more likely scenario to produce this system than two
back-to-back direct binary-binary interactions)
\citep{2004MNRAS.350..615G}.  \simon{Maybe add something about the
  curious binary AE Aurigae? - NL2: For sure.  I still don't know
  anything about it... But if I get time over the next few days I will
  do my research and write something.}

@@

\begin{figure}[ht!]
  \includegraphics[width=\columnwidth]{fig_orbital_evolution_for_12MSun.pdf}
  \includegraphics[width=\columnwidth]{fig_orbital_evolution_for_14MSun.pdf}
  \caption{Evolution of the total mass of the inner binary as a
    function of the orbital separation for four calculations with
    somewhat different initial conditions (see legend).  The initial
    binary for the orange curve is presented in
    fig.\,\ref{fig:topview_at_t0}, and in
    fig.\,\ref{fig:topview_at_t1000day} we present the final
    conditions.
\label{fig:mass_vs_semimajor_axis}}
\end{figure}
